Haploid Selection, Sex Ratio Bias, and Transitions Between sex-determination systems

### 2017-08-23 LaTeX/Turnover.tex
updated text with new figures

With haploid selection, we again find that transitions between male and female heterogamety can occur even if the new sex-determining region is less closely linked to the locus under selection, and when linkage is tight haploid selection in the heterogametic sex can cause strong sex ratio bias, which may increase or decrease with the spread of new sex chromsomes. 
These results indicate that favourable associations that develop between the ancestral sex-determining locus and selected loci can be broken during the spread of a new sex-determining region. 

@@

Overall, our models provide new predictions for the types of selection and the genomic location of loci that can drive transitions between sex-determination systems. %and for the factors determining the phylogenetic pattern of sex-determination systems. 

@@

Animals and angiosperms exhibit extremely diverse sex-determination systems \citep[reviewed in][]{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx}. 
Among species with genetic sex determination of diploid sexes, some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? 
mammals and most dioecious plants \citep{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining region changes. 
For example, transitions of the master sex-determining gene between chromosomes or the evolution of new master sex-determining genes have occurred in Salmonids \citep{Li:2011fm,Yano:2012di}, Diptera \citep{Vicoso:2015hf}, and \textit{Oryzias} \citep{Myosho:2012fv}.
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many gonochoric clades with genetic sex determination exhibit transitions between male (XY) and female (ZW) heterogamety, including snakes \textcolor{red}{(Gamble et al. 2017, Current Biology)}, lizards \citep{Ezaz:2009tk}, eight of 26 teleost fish families \citep{Mank:2006bt}, true fruit flies \citep[Tephritids,][]{Vicoso:2015hf}, amphibians \citep{Hillis:1990gu}, the angiosperm genus \textit{Silene} \citep{Slancarova:2013dq}, and Coleoptera and Hemiptera \citep[][plate 2]{Beukeboom:2014vb}.
Indeed, in some cases, both male and female heterogametic sex-determination systems can be found in the same species, as exhibited by some cichlid species \citep{Ser:2010iq} and \textit{Rana rugosa} \citep{Ogata:2007jm}.
In addition, multiple transitions have occurred between genetic and environmental sex-determination systems, e.g., in reptiles and fishes \citep{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc}.

@@

%\textcolor{red}{We add haploid selection:}
Here we extend current theory %on turnover between sex-determining systems 
by using mathematical models to find the conditions under which new sex-determination systems spread when individuals experience selection at both diploid and haploid stages. 
Even in animal and plant species that have much larger and more conspicuous diploid phases than haploid phases, many loci experience significant haploid selection through gamete competition and/or meiotic drive \citep{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-Mendelian) segregation of genotypes during gamete production (from one parent) and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool (potentially from multiple parents); the term `haploid selection' encompasses both processes. 
Meiotic drive generally occurs either during the production of male or female gametes only \citep{Ubeda:2005gw,Lindholm:2016cw}.
Because there are typically many more pollen/sperm than required for fertilization, gametic competition is also typically sex specific, occurring primarily among male gametes.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes are known to cause a response to selection \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}. 
Recent studies have demonstrated that sperm competition can alter haploid allele frequencies and increase offspring fitness \citep{Immler:2014im} \textcolor{red}{(Alavioon et al. 2017). 
Genetic mapping experiments, which are typically designed to minimize selection in diploids, have revealed segregation distortion in various species, including mice, Drosophila, Rice, Maize, Wheat, Barley, Cotton... 
In some of these cases, biased segregation has been shown to be attributable to meiotic drive and/or gametic selection \textcolor{red}{(Leppala et al. 2013, Didion et al. 2015, 2016 Xu et al 2013 (rice), Fishman...)}. }
% from Bai et al. (2016) ``[Segregation distortion] phenomenon is commonly detected via genetic mapping and has been documented in various species, including mouse (Eversley et al., 2010; Casellas et al., 2012), Drosophila (Phadnis and Orr, 2009; Larracuente and Presgraves, 2012; McDermott and Noor, 2012), Tigriopus (Pritchard et al., 2011), rice (Koide et al., 2012; Reflinur et al., 2014; Xu et al., 2014), maize (Tang et al., 2013), and cotton (Yu et al., 2011; Hulse-Kemp et al., 2015).'' Also more in that paper that might be useful. 
%Drosophila http://www.biorxiv.org/content/early/2017/02/06/106336

@@

Our models have two important new features. 
Firstly, when considering loci that are under selection and also in very tight linkage with the ancestral sex-determining region we explicitly calculate equilibrium allele frequencies.
This allows us to show that transitions between male and female heterogamety can evolve even when the neo-sex-determining locus is less closely linked to a locus under selection and therefore disrupts favourable ancestral associations between sex and the alleles selected in that sex.
%Such transitions are not favoured in models lacking tight linkage and/or haploid selection. 
Secondly, we allow sex-specific haploid selection to occur on a locus in tight or loose linkage with the ancestral sex-determining region. 
We find that sex-ratio biases caused by haploid selection can exert Fisherian sex-ratio selection upon novel sex-determiners but that their spread is also determined by selection on genetically-associated alleles. 
%Indeed, it is only when haploid-selected loci are tightly linked to the ancestral sex-determining region (and so sex-ratio biases are initially large) that we see an asymmetry between selection for XY to ZW transitions and ZW to XY transitions (e.g., because haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system).
Consequently, it is possible for selection on linked alleles to drive turnover between sex-determining systems despite causing transitory or even permanent increases in sex-ratio bias. 
%In addition, when linkage between the ancestral sex-determining locus and a locus under haploid selection is tight, sex ratio biases can be strong such that there is an asymmetry between invasion of ancestrally XY and ancestrally ZW systems because, e.g., haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system.

@@

%genetic system
We consider transitions between ancestral and novel sex-determining systems using a three-locus model, each locus having two alleles. 
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X}; $XX$ genotypes become females and $XY$ become males (or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is epistatically dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, we can interpret $m$ as an environmental sex determination (ESD) allele, such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
We also analyze a model of maternally-controlled environmental sex-determination, where mothers with at least one $m$ allele produce daughters with probability $k$.

@@

The evolution of a new sex-determination system requires that a rare mutant allele at the novel sex-determining locus, $m$, increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system of eight equations describing the frequency of eggs and sperm carrying the $m$ allele in the next generation (equations \ref{eq:recursions}). %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
This system simplifies substantially in a number of cases of interest. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males or by females, respectively. 
Furthermore, if the $m$ allele is fully epistatically dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Thus, the invasion of rare dominant neo-Y or neo-W alleles is determined by the largest eigenvalue that solves a quadratic characteristic polynomial,
$\lambda^2+ b \lambda + c =0$.
%\begin{equation}
%\lambda^2+ b \lambda + c =0
%\label{eq:charpoly_neoY}
%\end{equation}
%
%\noindent
%where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}+\lambda_{ma})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\rho_{mA}\rho_{ma}$ (see table \ref{tab:haplotype_growth}).
%Here, $b= - (\lambda_{mA} + \lambda_{ma})$ and $c = \lambda_{mA} \lambda_{ma} -\rho_{mA} \rho_{ma}$, where $\lambda_{mi}$ is the (multiplicative) growth rate of mutant haplotypes on background $i\in\{A,a\}$, accounting for loss due to recombination, and $\rho_{mi}$ is the rate of addition of mutant haplotypes onto background $i\in\{A,a\}$ due to recombination (see table \ref{tab:haplotype_growth}).
Here, $b= - (\lambda_{mA} + \lambda_{ma})+(\rho_{mA} + \rho_{ma})$ and $c = (\lambda_{mA}-\rho_{mA}) (\lambda_{ma}-\rho_{ma}) -\rho_{mA} \rho_{ma}$, where $\lambda_{mi}$ is the multiplicative growth rate of mutant haplotypes on background $i\in\{A,a\}$, without accounting for loss due to recombination, and $\rho_{mi}$ is the rate at which mutant haplotypes on background $i\in\{A,a\}$ recombine onto the other \textbf{A} locus background in heterozygotes (see Table \ref{tab:haplotype_growth}).
The $\lambda_{mi}$ and $\rho_{mi}$, and thus the spread of the mutant $m$ allele, depend on the frequency of alleles at the \textbf{A} and \textbf{X} loci in the ancestral population. 
In the ancestral population, it is convenient to follow the frequency of the $A$ allele among female gametes (eggs), $p^\female_X$, and among X-bearing, $p^\male_X$, and among Y-bearing, $p^\male_Y$, male gametes (sperm/pollen). 
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males.

@@

The equilibrium allele frequencies and their stability conditions are given in the appendix, when there is complete linkage between the ancestral sex-determining region and the \textbf{A} locus ($r=0$). Here, either the $A$ allele or the $a$ allele must be fixed on the Y. 
% because selection on the Y is effectively haploid an incapably of maintaining a polymorphism in the absence of frequency dependent selection. 
Because the labelling of alleles is arbitrary, we will assume that the $a$ locus is fixed on the Y ($p^\male_Y=0$), without loss of generality . 
If there are two alleles maintained at the \textbf{A} locus, the X can either be fixed for the $A$ allele ($p^\female_X=p^\male_X=1$) or polymorphic (0<$p^\female_X, p^\male_X$<1). 
%However, because the X is present in both males and females, there are distinct selection pressures on X's in females and X's in males, which will always be paired with an $A$ allele from the Y.

@@

\textcolor{red}{Sally edits only looked at up to this point. Next task: figures to match with this tight linkage section.}

%\textcolor{blue}{
%Briefly, the neo-W invades under the ancestral XY sex-determining system, the X chromosome may not be able to as specialised on a female-beneficial allele
%The full characteristic polynomials are given in the appendix (equations \ref{Binvasion} and \ref{Ainvasion}).
Neo-W alleles, on the other hand, can invade an ancestral XY system under some conditions (the full invasion conditions are given in the appendix; equations \ref{Binvasion} and \ref{Ainvasion}). 
Significantly, we note that it is possible for both neo-W haplotypes to spread ($\lambda_{mA}>1$ and $\lambda_{ma}>1$), in which case neo-W invasion can occur regardless of linkage to the selected locus.
That is, selection on loci within the non-recombining region of the SDR can favour the invasion of a less closely linked neo-W (e.g., on an autosome). 
Although haploid selection can favour neo-W alleles because the ancestral sex ratio becomes male biased, this is not the only circumstance in which less tightly linked neo-W alleles invade. 
For example, unlinked neo-W alleles can invade in the absence of any haploid selection. 
This result is unexpected given the results of \cite{vanDoorn:2010hu}, who did not explicitly calculate equilibrium allele frequencies under tight linkage and generally concluded that heterogametic transitions occur when neo-sex-determining alleles are in tighter linkage with loci under sex-specific diploid selection. 
Therefore, we focus on cases where there is no haploid selection and discuss the effects of haploid selection in the appendix. 
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Tighter linkage allows looser sex-linkage to evolve, even in absence of haploid selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\centerline{\includegraphics[width=1.5\linewidth]{PositionPlot_SexAntagTighter_Mike}}
\caption{
Transitions between XY and ZW systems can occur even when the neo-SDR is more loosely linked to a locus under sexually-antagonistic selection (here, without haploid selection $t^\Hermaphrodite = \alpha^\Hermaphrodite_\Delta = 0$).
In panel A, linkage is loose enough relative to selection that the weak selection analytical results hold, and a neo-W can only invade when it is more tightly linked with the selected locus ($R<r$; shaded region).
In panel B, linkage is tight enough relative to selection that the weak selection analytical results do no hold, and a neo-W can only invade even when it is less tightly linked with the selected locus ($r<R$; unshaded region).
In panel C we vary the recombination rate between the neo-W and the selected locus ($R$) for a fixed recombination rate between the ancestral-SDR and the selected locus ($r=0.005$). 
Coloured markers show recombination rates for which the temporal dynamics of invasion are plotted in the inset, demonstrating that neo-W alleles can fix (reach frequency $0.5$ among female gametes) if they are more (black) or less (red) closely linked to a locus experiencing sexually-antagonistic selection. 
A very loosely linked neo-W does not spread in this case (blue and green lines overlap and go to 0). 
Fitness parameters are shown by an asterisk in Figure \ref{fig:regionplots}: $w_{AA}^\female = 1.05$, $w_{aa}^\male = 1.2$, $w_{aa}^\female = w_{AA}^\male = 0.85$, $w_{Aa}^\Hermaphrodite = 1$.
}
\label{fig:SexAntagTighter}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\textcolor{blue}{
If we categorise the $a$ allele as being ancestrally `male-beneficial' via the fact that it is fixed on the Y, then $\lambda_{mA}>1$ indicates that the neo-W spreads when found with the ancestrally `female-beneficial' allele. 
Intuitively, this is possible because the ancestral X chromosome is not able to specialise on the `female-beneficial' allele due to the fact that X's are sometimes found in males. 
For example, a polymorphism of $A$ and $a$ alleles can be maintained on the X despite directional selection in favour of the $A$ allele in females ($s^\female>0$, $0<h^\female<1$) because the $a$ allele is favoured in males. 
W-$A$ haplotypes will only create females with high fitness $AA$ or $Aa$ genotypes and can therefore have higher fitness than ancestral females, which sometimes also produce $aa$ females. 
Thus, the neo-W can spread by allowing increased specialization on female beneficial alleles.
%As another example, $AA$ individuals can have low fitness in females and yet the $A$ can be fixed on the X due to strong overdominance in males favouring X$A$Y$a$ males. 
%In this case, W$A$ haplotypes produce fewer unfit $AA$ females because they produce some females through matings with Y$a$-bearing sperm. 
%}

>>>>

With haploid selection, we again find that transitions between male and female heterogamety can occur even if the new sex-determining region is less closely linked to the locus under selection. 
Haploid selection in the heterogametic sex can also cause sex ratio biases, which may increase or decrease with the spread of new sex chromosomes. 
Thus, transitions between sex-determination systems cannot be simply predicted by selection to equalise the sex-ratio. 
%These results indicate that loci experiencing haploid selection can favour the evolution of new sex-determination systems. 
%These results indicate that favourable associations that develop between the ancestral sex-determining locus and selected loci can be broken during the spread of a new sex-determining region. 

@@

Overall, our models reveal that transitions between sex-determination systems, particularly transitions where the heterogametic sex changes, can be driven by loci in unexpected genomic locations that experience selection during diploid and/or haploid phases, which might be reflected by the lability with which sex-determination systems evolve. %and for the factors determining the phylogenetic pattern of sex-determination systems. 

@@

Animals and angiosperms exhibit extremely diverse sex-determination systems \citep[reviewed in][]{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx}. 
Among species with genetic sex determination of diploid sexes, some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? 
mammals and most dioecious plants \citep{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining region changes. 
For example, transitions of the master sex-determining gene between chromosomes or the evolution of new master sex-determining genes have occurred in Salmonids \citep{Li:2011fm,Yano:2012di}, Diptera \citep{Vicoso:2015hf}, and \textit{Oryzias} \citep{Myosho:2012fv}.
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many gonochoric clades with genetic sex determination exhibit transitions between male (XY) and female (ZW) heterogamety, including snakes \textcolor{red}{(Gamble et al. 2017, Current Biology)}, lizards \citep{Ezaz:2009tk}, eight of 26 teleost fish families \citep{Mank:2006bt}, true fruit flies \citep[Tephritids,][]{Vicoso:2015hf}, amphibians \citep{Hillis:1990gu}, the angiosperm genus \textit{Silene} \citep{Slancarova:2013dq}, \textcolor{red}{the angiosperm family \textit{Salicaceae} (Pucholt et al. 2015, 2017),} and Coleoptera and Hemiptera \citep[][plate 2]{Beukeboom:2014vb}.
Indeed, in some cases, both male and female heterogametic sex-determination systems can be found in the same species, as exhibited by some cichlid species \citep{Ser:2010iq} and \textit{Rana rugosa} \citep{Ogata:2007jm} \textcolor{red}{(Miura 2007)}. 
In addition, multiple transitions have occurred between genetic and environmental sex-determination systems, e.g., in reptiles and fishes \citep{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc}.

@@

%\textcolor{red}{We add haploid selection:}
Here we use mathematical models to find the conditions under which new sex-determination systems spread when individuals experience selection at both diploid and haploid stages. 
Even in animal and plant species that have much larger and more conspicuous diploid phases than haploid phases, many loci experience significant haploid selection through gamete competition and/or meiotic drive \citep{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-Mendelian) segregation of genotypes during gamete production (from one parent) and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool (potentially from multiple parents); the term `haploid selection' encompasses both processes. 
\textcolor{red}{
Genetic mapping experiments, which are typically designed to minimize selection in diploids, have revealed segregation distortion in various species, including mice, Drosophila, Rice, Maize, Wheat, Barley, Cotton... 
In some of these cases, biased segregation has been attributed to meiotic drive and/or gametic selection \textcolor{red}{(Leppala et al. 2013, Didion et al. 2015, 2016 Xu et al 2013 (rice), Fishman...)}. }
% from Bai et al. (2016) ``[Segregation distortion] phenomenon is commonly detected via genetic mapping and has been documented in various species, including mouse (Eversley et al., 2010; Casellas et al., 2012), Drosophila (Phadnis and Orr, 2009; Larracuente and Presgraves, 2012; McDermott and Noor, 2012), Tigriopus (Pritchard et al., 2011), rice (Koide et al., 2012; Reflinur et al., 2014; Xu et al., 2014), maize (Tang et al., 2013), and cotton (Yu et al., 2011; Hulse-Kemp et al., 2015).'' Also more in that paper that might be useful. 
%Drosophila http://www.biorxiv.org/content/early/2017/02/06/106336
Meiotic drive generally occurs either during the production of male or female gametes only \citep{Ubeda:2005gw,Lindholm:2016cw}.
Because there are typically many more pollen/sperm than required for fertilization, gametic competition is also typically sex specific, occurring primarily among male gametes.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes are known to cause a response to selection \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}. 
Recent studies have demonstrated that sperm competition can alter haploid allele frequencies and increase offspring fitness \citep{Immler:2014im} \textcolor{red}{(Alavioon et al. 2017)}.

@@

%Perhaps longer discussion of tight sex-linkage here? e.g., mention that r=0 is a particularly significant case for sex chromosomes.

We find that sex-ratio biases caused by haploid selection can exert Fisherian sex-ratio selection upon novel sex-determiners but that their spread is also determined by selection on genetically-associated alleles. 
%Indeed, it is only when haploid-selected loci are tightly linked to the ancestral sex-determining region (and so sex-ratio biases are initially large) that we see an asymmetry between selection for XY to ZW transitions and ZW to XY transitions (e.g., because haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system).
Consequently, it is possible for selection on linked alleles to drive turnover between sex-determining systems despite causing transitory or even permanent increases in sex-ratio bias. 
Another significant feature of our model is that we consider selection loci that are under selection and also in very tight linkage with the ancestral sex-determining region.
Even in the absence of haploid selection, we show that transitions between male and female heterogamety can then evolve despite the fact that the neo-sex-determining locus is less closely linked to a locus under selection and therefore disrupts favourable ancestral associations between sex and the alleles selected in that sex.
%Such transitions are not favoured in models lacking tight linkage and/or haploid selection. 
%Secondly, we allow sex-specific haploid selection to occur on a locus in tight or loose linkage with the ancestral sex-determining region. 
%In addition, when linkage between the ancestral sex-determining locus and a locus under haploid selection is tight, sex ratio biases can be strong such that there is an asymmetry between invasion of ancestrally XY and ancestrally ZW systems because, e.g., haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system.

@@

\textcolor{red}{switch between $\chi$ and $\rho$ in all places because $\chi$ is used for double recombination events.}

\textcolor{red}{Change $\zeta$ to represent zygotic sex ratio of males, consistent with $q$ and figures.}

%genetic system
We consider transitions between ancestral and novel sex-determining systems using a three-locus model, each locus having two alleles. 
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X}; $XX$ genotypes become females and $XY$ become males (or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is epistatically dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, we can interpret $m$ as an environmental sex determination (ESD) allele, such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
%We also analyze a model of maternally-controlled environmental sex-determination, where mothers with at least one $m$ allele produce daughters with probability $k$.

@@

The evolution of a new sex-determination system requires that a rare mutant allele at the novel sex-determining locus, $m$, increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system of eight equations describing the frequency of eggs and sperm carrying the $m$ allele in the next generation (equations \ref{eq:recursions}). %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
This system simplifies substantially in a number of cases of interest. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males or by females, respectively. 
Furthermore, if the $m$ allele is fully epistatically dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Thus, the invasion of rare dominant neo-Y or neo-W alleles is determined by the largest eigenvalue that solves a quadratic characteristic polynomial,
\textcolor{red}{Mention the possibility that the other roots yield the leading eigenvalue somewhere.}
$\lambda^2+ b \lambda + c =0$.
%\begin{equation}
%\lambda^2+ b \lambda + c =0
%\label{eq:charpoly_neoY}
%\end{equation}
%
%\noindent
%where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}+\lambda_{ma})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\rho_{mA}\rho_{ma}$ (see table \ref{tab:haplotype_growth}).
%Here, $b= - (\lambda_{mA} + \lambda_{ma})$ and $c = \lambda_{mA} \lambda_{ma} -\rho_{mA} \rho_{ma}$, where $\lambda_{mi}$ is the (multiplicative) growth rate of mutant haplotypes on background $i\in\{A,a\}$, accounting for loss due to recombination, and $\rho_{mi}$ is the rate of addition of mutant haplotypes onto background $i\in\{A,a\}$ due to recombination (see table \ref{tab:haplotype_growth}).
Here, $b= - (\lambda_{mA} + \lambda_{ma})+(\rho_{mA} + \rho_{ma})$ and $c = (\lambda_{mA}-\rho_{mA}) (\lambda_{ma}-\rho_{ma}) -\rho_{mA} \rho_{ma}$, where $\lambda_{mi}$ is the multiplicative growth rate of mutant haplotypes on background $i\in\{A,a\}$, without accounting for loss due to recombination, and $\rho_{mi}$ is the rate at which mutant haplotypes on background $i\in\{A,a\}$ recombine onto the other \textbf{A} locus background in heterozygotes (see Table \ref{tab:haplotype_growth}).
The $\lambda_{mi}$ and $\rho_{mi}$, and thus the spread of the mutant $m$ allele, depend on the frequency of alleles at the \textbf{A} and \textbf{X} loci in the ancestral population. 
In the ancestral population, it is convenient to follow the frequency of the $A$ allele among female gametes (eggs), $p^\female_X$, and among X-bearing, $p^\male_X$, and among Y-bearing, $p^\male_Y$, male gametes (sperm/pollen). 
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males.

@@

The ancestral equilibrium allele frequencies and their stability conditions are given in the appendix.
When there is complete linkage between the ancestral sex-determining region and the \textbf{A} locus ($r=0$), either the $A$ allele or the $a$ allele must be fixed on the Y. 
% because selection on the Y is effectively haploid an incapably of maintaining a polymorphism in the absence of frequency dependent selection. 
Because the labelling of alleles is arbitrary, we will assume that the $a$ locus is fixed on the Y ($p^\male_Y=0$), without loss of generality. 
If there are two alleles maintained at the \textbf{A} locus, the X can either be fixed for the $A$ allele ($\hat{p}^\female_X=\hat{p}^\male_X=1$) or polymorphic (0<$\hat{p}^\female_X, \hat{p}^\male_X$<1). 
%However, because the X is present in both males and females, there are distinct selection pressures on X's in females and X's in males, which will always be paired with an $A$ allele from the Y.

@@

%\textcolor{blue}{
Neo-W alleles, on the other hand, can invade an ancestral XY system under some conditions (the full invasion conditions are given in the appendix; equations \ref{Binvasion} and \ref{Ainvasion}). 
That is, selection on loci within the non-recombining region of the SDR can favour the invasion of a less closely linked neo-W, see Figure \ref{fig:SexAntagTighter}. 
%Although haploid selection can favour neo-W alleles because the ancestral sex ratio becomes male biased, this is not the only circumstance in which less tightly linked neo-W alleles invade. 
%\textcolor{red}{Sally: unpack, give one condition and relate to figures}
%For example, we note that it is possible for both neo-W haplotypes to spread ($\lambda_{mA}>1$ and $\lambda_{ma}>1$), in which case neo-W invasion can occur regardless of linkage to the selected locus. 
%For example, unlinked neo-W alleles can invade in the absence of any haploid selection. 
This result is unexpected given the results of \cite{vanDoorn:2010hu}, who did not explicitly calculate equilibrium allele frequencies under tight linkage and generally concluded that heterogametic transitions occur when neo-sex-determining alleles are in tighter linkage with loci under sex-specific diploid selection. 
To develop an understanding (intuition) for how this happens, we focus on cases where there is no haploid selection and discuss the effects of haploid selection in the appendix. 
%}
%Briefly, a neo-W can spread when the females it produces have higher fitness than the females produced under the ancestral SDR, which all have genotype XX. X chromosomes are sometimes found in males and therefore allele frequencies on the X can be suboptimal for females. In addition, the neo-W produces females from matings with Y-bearing male gametes. The allele associated with the Y can also be beneficial in females. 
%the neo-W invades under the ancestral XY sex-determining system, the X chromosome may not be able to as specialised on a female-beneficial allele

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Tighter linkage allows looser sex-linkage to evolve, even in absence of haploid selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\centerline{\includegraphics[width=1.5\linewidth]{PositionPlot_SexAntagTighter_Mike}}
\caption{
Transitions between XY and ZW systems can occur even when the neo-SDR is more loosely linked to a locus under sexually-antagonistic selection (here, without haploid selection $t^\Hermaphrodite = \alpha^\Hermaphrodite_\Delta = 0$).
In panel A, linkage is loose enough relative to selection that the weak selection analytical results hold, and a neo-W can only invade when it is more tightly linked with the selected locus ($R<r$; shaded region).
In panel B, linkage is tight enough relative to selection that the weak selection analytical results do not hold, and a neo-W can only invade even when it is less tightly linked with the selected locus ($r<R$; unshaded region).
In panel C we vary the recombination rate between the neo-W and the selected locus ($R$) for a fixed recombination rate between the ancestral-SDR and the selected locus ($r=0.005$). 
Coloured markers show recombination rates for which the temporal dynamics of invasion are plotted in the inset, demonstrating that neo-W alleles can fix (reach frequency $0.5$ among female gametes) if they are more (black) or less (red) closely linked to a locus experiencing sexually-antagonistic selection. 
A very loosely linked neo-W does not spread in this case (blue and green lines overlap and go to 0). Indeed, we show that neo-W invasion fitness is negative when $R=1/2$ and $s^\female s^\male<0$, $0<h^\Hermaphrodite<1$) in the supplementary \textit{Mathematica} file. 
Fitness parameters are shown by an asterisk in Figure \ref{fig:regionplots}: $w_{AA}^\female = 1.05$, $w_{aa}^\male = 1.2$, $w_{aa}^\female = w_{AA}^\male = 0.85$, $w_{Aa}^\Hermaphrodite = 1$.
\textcolor{red}{consider removing panel A, which is repeated in Figure 3.}
}
\label{fig:SexAntagTighter}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
